Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain

### 2018-10-15 pkg/disttree/inst/draft/draft.Rnw
some more revision based on reviews and feedback
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2326 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 0f09150666914e1d76d6048a151708759a87441a

\Abstract{
To obtain a probabilistic model for a dependent variable based on some set of
explanatory variables, a distributional approach is often adopted where the
parameters of the distribution are linked to regressors. In many classical
models this only captures the location of the distribution but
over the last decade there has been increasing interest in distributional
regression approaches modeling all parameters including location, scale, and
shape. Notably, so-called non-homogenous Gaussian regression (NGR) models both
mean and variance of a Gaussian response and is particularly
popular in weather forecasting. More generally, the GAMLSS framework
allows to establish generalized additive models for location, scale, and shape
with smooth linear or nonlinear effects.
%
However, when variable selection is required and/or there are
non-smooth dependencies or interactions (especially unknown or of high-order),
it is challenging to establish a good GAMLSS. A natural alternative in these
situations would be the application of regression trees or random forests but,
so far, no general distributional framework is available for these. Therefore,
a framework for distributional regression trees and forests is proposed that blends
regression trees and random forests with classical distributions from the GAMLSS
framework as well as their censored or truncated counterparts.
%
To illustrate these novel approaches in practice, they are employed to obtain
probabilistic precipitation forecasts at numerous sites in a mountainous
region (Tyrol, Austria) based on a large number of numerical weather
prediction quantities. It is shown that the novel distributional regression forests
automatically select variables and interactions, performing on par or often even
better than GAMLSS specified either through prior meteorological knowledge or a
computationally more demanding boosting approach.
}
\Keywords{parametric models, regression trees, random forests, recursive partitioning, probabilistic forecasting, GAMLSS}

@@

# add labels
text(x = -1.7, y = lh1, labels = "2009", col = pal[1], cex = 0.8)     # -0.8
text(x = -1.7, y = lh2, labels = "2010", col = pal[3], cex = 0.8)     # -0.8
text(x = -1.7, y = lh3, labels = "2011", col = pal[2], cex = 0.8)     # -0.8
text(x = -1.7, y = lh4, labels = "2012", col = pal[4], cex = 0.8)     # -0.8
@
\caption{\label{fig:axams-dist}\hlc{Total precipitation predictions by a 
distributional forest at station Axams for July 24 in 2009, 2010, 2011 and 
2012 learned on data from 1985--2008. Observations are non-negative and 
modeled by a Gaussian distribution, left-censored at 0.
The observations are depicted by crosses and the predicted point mass
from the model by filled circles.}}
\end{figure}

In weather forecasting, these properties are especially appealing in 
mountainous regions and complex terrain where a wide range of local-scale
effects are not yet resolved by the NWP models. 
Thus, effects with abrupt changes and possibly nonlinear interactions 
might be required to account for site-specific unresolved features.
To illustrate this in practice,
precipitation forecasts are obtained with distributional forests at 95
meteorological stations in a mountainous region in the Alps, covering mainly
Tyrol, Austria, and adjacent areas (see the map in Figure~\ref{fig:all-map}).
More specifically, a Gaussian distribution, left-censored at zero, is employed to model 24-hour total precipitation so that the
zero-censored point mass describes the probability of observing no precipitation 
on a given day (see Figure~\ref{fig:axams-dist}). Forecasts for July are
established based on data from the same month over the years 1985--2012 including
80~covariates derived from a wide range of different NWP quantities. As
Figure~\ref{fig:axams-dist} shows, the station-wise forests yield a full
distributional forecast for each day -- here for one specific day (July~24) at one
station (Axams) over four years (2009--2012) -- based on the previous 24~years
as learning data. The corresponding observations conform reasonably well with
the predictions. In Section~\ref{sec:precipitation} we investigate the
performance of distributional forests in this forecasting task in more detail.
Compared to three alternative zero-censored Gaussian models distributional forests
perform at least on par and sometimes clearly better while requiring no meteorological
knowledge about the athmospheric processes which drive formation of precipitation
for the model specification. The three alternatives are: a standard ensemble model
output statistics approach \citep[EMOS,][]{Gneiting+Raftery+Westveld:2005} based
on an NGR, a GAMLSS with regressors prespecified based on meteorological expertise
\citep[following][]{Stauffer+Umlauf+Messner:2017}, and a boosted GAMLSS
\citep{Hofner+Mayr+Schmid:2016} using non-homogeneous boosting
\citep{Messner+Mayr+Zeileis:2017} as an alternative technique for variable selection among all
80~available regressors.

@@

The general distributional notation is exemplified in all three steps for the
Gaussian distribution, left-censored at zero. The latter is employed in the empirical 
case study in Section~\ref{sec:precipitation} to model power-transformed daily 
precipitation amounts.

@@

Many statistical weather forecasting models leverage the strengths of modern
numerical weather prediction (NWP) systems \citep[see][]{Bauer+Thorpe+Brunet:2015}.  
One frequently used method based on distributional regression models is the 
ensemble model output statistics (EMOS) approach first proposed by
\cite{Gneiting+Raftery+Westveld:2005} to produce high-quality forecasts for
specific quantities and sites.  
\hlc{In case of precipitation forecasting EMOS typically uses the ensemble mean of 
``total precipitation'' (\emph{tp}) forecasts as predictor for the location 
parameter $\mu$ and the corresponding ensemble standard deviation for the 
scale parameter $\sigma$ of the statistical model based on the assumption
of the response following a left-censored at zero Gaussian distribution.
This distributional approach of modeling both parameters allows to correct 
for possible errors of the ensemble in both, the expectation but also the 
uncertainty of a specific forecast. In order to assure positivity of the 
scale parameter a log-link is applied. Hence, the two distribution parameters 
are modeled by the linear functions} 
$\mu = \beta_0 + \beta_1 \cdot \mbox{mean}(\mbox{tp}^{\frac{1}{p}})$ and  
$\log(\sigma) = \gamma_0 + \gamma_1 \cdot \mbox{sd}(\mbox{tp}^{\frac{1}{p}})$
with power transformation parameter $\frac{1}{p}$ and 
regression coefficients $\beta_0$, $\beta_1$, $\gamma_0$ and $\gamma_1$.  
\hlc{(Hereafter, this model is going to be referred to as a basic EMOS model.)}

@@

In the following, it is illustrated how distributional forests can
solve this problem without requiring prior expert knowledge.  For fitting the
forest only the response distribution and the list of potential predictor
variables need to be specified (along with a few algorithmic details) and then
the relevant variables, interactions, and potentially nonlinear effects are
determined automatically in a data-driven way. Here, we employ a zero-censored
Gaussian distribution and 80~predictor variables computed from ensemble means
and spreads of various NWP outputs. The predictive performance of the forest is
compared to three other zero-censored Gaussian models: (a)~a standard basic
EMOS, (b)~a GAMLSS with prespecified effects and interactions based on
meteorological knowledge/experience, and (c)~a boosted GAMLSS with automatic
selection of smooth additive terms based on all 80~predictor variables.

@@

\item \emph{EMOS:} Standard ensemble model output statistics
  models use the ensemble mean of total precipitation as regressor in the location
  submodel and the corresponding ensemble standard deviation in the scale submodel.
  The parameters are estimated by maximum likelihood, using an identity
  link for the location part and a log link for the scale part
  \citep[following the advice of][]{Gebetsberger+Messner+Mayr:2017}.

@@

\subsubsection*{East Tyrol}
Figure~\ref{fig:all-map} \hlc{shows that, contrary to the rest of Tyrol, 
for the majority of stations in East Tyrol the prespecified GAMLSS performs
best in this particular setting (learning data: 1985--2008, testing data:
2009--2012). East Tyrol is located south of the main Alpine Ridge and 
exhibits different climatological characteristics than the part north of 
the Alpine Ridge. It seems that the choice of covariates and their interactions 
we made for the prespecified GAMLSS method is particularly adequate for the 
region of East Tyrol for this specific analysis. However, the advantages of 
the prespecified GAMLSS models in this particular setting (training: 1985--2008, 
validation: 2009--2012) even out in a cross-validation framework for most of 
the stations in East Tyrol. These results are presented in Supplement 2 
(Stationwise Evaluation). There, the exact same evaluations as in} 
Section~\ref{sec:axams} \hlc{for station Axams have been carried out 
for 14 other stations.}

@@

\subsubsection{Distributional specifications}
%% choice of distribution mit Verweis auf Supplement distributions
\hlc{Choosing an adequate distributional family is a crucial step and forms
the basis required to build a well fitting model. Therefore, this
decision should be made thoughtfuly. 
In the presented application we applied a Gaussian left-censored at
zero distribution as this has proven to be a well working choice in
earlier literature} \citep{Stauffer+Mayr+Messner:2017}. 
\hlc{But to test for robustness against distributional misspecifation 
different distributional assumptions have been considered as well. 
In Supplement 1 (Different Response Distributions) the exact
same evaluations as in} Section~\ref{sec:all} \hlc{are carried out for 15
observation stations, once employing a Gaussian hurdle model and once
with a left-censored at zero logistic distribution in order to account
for heavier tails. Under both assumptions the obtained results 
are similar to those presented in the paper applying a left-censored
at zero Gaussian distribution. However, comparing the two censored distributions,
at many stations the models perform not as well under the logisctic distribution 
as under the Gaussian distribution. The hurdle model leads to slight improvements 
in some settings, but this comes with more variablity (especially for EMOS and the 
boosted GAMLSS).
Alltogether, the distributional forest has proven to be very robust against
distributional misspecifications.}

\subsubsection*{Computational difficulties}
%% GAMLSS errors with references gamboostLSS Hofner et al.
As stated by \cite{Hofner+Mayr+Schmid:2016} \hlc{the AIC-based variable selection 
methods implemented in the} \textsf{R} package \textbf{gamlss} \hlc{``can be unstable, 
especially when it comes to selecting possibly different sets of variables for 
multiple distribution parameters.'' We have noticed computational problems when
applying the function} \textbf{gamlss} \hlc{in certain settings within the cross-validation
framework as it did not succeed in fitting a model. In this cases the method
was not taken into consideration in the comparison of all applied models.}

@@

\section{PIT histograms}
\label{app:pit}

\begin{figure}[b!]
\centering
%\begin{subfigure}{0.8\textwidth}
<<rain_axams_pit_oosample, fig=TRUE, echo=FALSE, width = 7>>=

@@

@
%\end{subfigure}
\caption{\label{fig:axams-pit}Out-of-sample PIT histograms (2009--2012) for station Axams and models learned on data from 1985--2008.}
\end{figure}

As an alternative to QQ~plots based on randomized quantile residuals,
probability integral transform (PIT, \citealt{Gneiting+Balabdaoui+Raftery:2007}) 
histograms are a commonly used tool to assess how well-calibrated distributional 
fits are. Figure~\ref{fig:axams-pit} shows out-of-sample PIT histograms for the same
probabilistic forecasts as Figure~\ref{fig:axams-qq}, confirming that despite
some deviations the overall calibration of all models is acceptable.

>>>>

\Abstract{
To obtain a probabilistic model for a dependent variable based on some set of
explanatory variables, a distributional approach is often adopted where the
parameters of the distribution are linked to regressors. In many classical
models this only captures the location of the distribution but
over the last decade there has been increasing interest in distributional
regression approaches modeling all parameters including location, scale, and
shape. Notably, so-called non-homogenous Gaussian regression (NGR) models both
mean and variance of a Gaussian response and is particularly
popular in weather forecasting. 
% More generally, the GAMLSS framework allows to establish generalized additive 
% models for location, scale, and shape with smooth linear or nonlinear effects.
Less specifically, generalized additive models for location, scale, and shape (GAMLSS) provide a framework where each distribution parameter is modeled separately capturing smooth linear or nonlinear effects.
%
However, when variable selection is required and/or there are
non-smooth dependencies or interactions (especially unknown or of high-order),
it is challenging to establish a good GAMLSS. A natural alternative in these
situations would be the application of regression trees or random forests but,
so far, no general distributional framework is available for these. Therefore,
a framework for distributional regression trees and forests is proposed that blends
regression trees and random forests with classical distributions from the GAMLSS
framework as well as their censored or truncated counterparts.
%
To illustrate these novel approaches in practice, they are employed to obtain
probabilistic precipitation forecasts at numerous sites in a mountainous
region (Tyrol, Austria) based on a large number of numerical weather
prediction quantities. It is shown that the novel distributional regression forests
automatically select variables and interactions, performing on par or often even
better than GAMLSS specified either through prior meteorological knowledge or a
computationally more demanding boosting approach.
}
\Keywords{parametric models, regression trees, random forests, recursive partitioning, probabilistic forecasting, GAMLSS}

@@

# add labels
text(x = -1.7, y = lh1, labels = "2009", col = pal[1], cex = 0.8)     # -0.8
text(x = -1.7, y = lh2, labels = "2010", col = pal[3], cex = 0.8)     # -0.8
text(x = -1.7, y = lh3, labels = "2011", col = pal[2], cex = 0.8)     # -0.8
text(x = -1.7, y = lh4, labels = "2012", col = pal[4], cex = 0.8)     # -0.8
@
\caption{\label{fig:axams-dist}\hlc{Total precipitation predictions by a 
distributional forest at station Axams for July 24 in 2009, 2010, 2011 and 
2012 learned on data from 1985--2008. Observations are non-negative and 
modeled by a Gaussian distribution left-censored at zero.
The observations are depicted by crosses and the predicted point mass
from the model by filled circles.}}
\end{figure}

In weather forecasting, these properties are especially appealing in 
mountainous regions and complex terrain where a wide range of local-scale
effects are not yet resolved by the NWP models. 
Thus, effects with abrupt changes and possibly nonlinear interactions 
might be required to account for site-specific unresolved features.
To illustrate this in practice,
precipitation forecasts are obtained with distributional forests at 95
meteorological stations in a mountainous region in the Alps, covering mainly
Tyrol, Austria, and adjacent areas (see the map in Figure~\ref{fig:all-map}).
More specifically, a Gaussian distribution left-censored at zero, is employed to model 24-hour total precipitation so that the
zero-censored point mass describes the probability of observing no precipitation 
on a given day (see Figure~\ref{fig:axams-dist}). Forecasts for July are
established based on data from the same month over the years 1985--2012 including
80~covariates derived from a wide range of different NWP quantities. As
Figure~\ref{fig:axams-dist} shows, the station-wise forests yield a full
distributional forecast for each day -- here for one specific day (July~24) at one
station (Axams) over four years (2009--2012) -- based on the previous 24~years
as learning data. The corresponding observations conform reasonably well with
the predictions. In Section~\ref{sec:precipitation} we investigate the
performance of distributional forests in this forecasting task in more detail.
Compared to three alternative zero-censored Gaussian models distributional forests
perform at least on par and sometimes clearly better while requiring no meteorological
knowledge about the athmospheric processes which drive formation of precipitation
for the model specification. The three alternatives are: a standard ensemble model
output statistics approach \citep[EMOS,][]{Gneiting+Raftery+Westveld:2005} based
on an NGR, a GAMLSS with regressors prespecified based on meteorological expertise
\citep[following][]{Stauffer+Umlauf+Messner:2017}, and a boosted GAMLSS
\citep{Hofner+Mayr+Schmid:2016} using non-homogeneous boosting
\citep{Messner+Mayr+Zeileis:2017} as an alternative technique for variable selection among all
80~available regressors.

@@

The general distributional notation is exemplified in all three steps for the
Gaussian distribution left-censored at zero. The latter is employed in the empirical 
case study in Section~\ref{sec:precipitation} to model power-transformed daily 
precipitation amounts.

@@

Many statistical weather forecasting models leverage the strengths of modern
numerical weather prediction (NWP) systems \citep[see][]{Bauer+Thorpe+Brunet:2015}.  
One frequently used method based on distributional regression models is the 
ensemble model output statistics (EMOS) approach first proposed by
\cite{Gneiting+Raftery+Westveld:2005} to produce high-quality forecasts for
specific quantities and sites.  
\hlc{In case of precipitation forecasting EMOS typically uses the ensemble mean of 
``total precipitation'' (\emph{tp}) forecasts as predictor for the location 
parameter $\mu$ and the corresponding ensemble standard deviation for the 
scale parameter $\sigma$ of the statistical model based on the assumption
of the response following a Gaussian distribution left-censored at zero.
This distributional approach of modeling both parameters allows to correct 
for possible errors of the NWP ensemble in both, the expectation but also the 
uncertainty of a specific forecast. In order to assure positivity of the 
scale parameter a log-link is applied. Hence, the two distribution parameters 
are modeled by the linear functions} 
$\mu = \beta_0 + \beta_1 \cdot \mbox{mean}(\mbox{tp}^{\frac{1}{p}})$ and  
$\log(\sigma) = \gamma_0 + \gamma_1 \cdot \mbox{sd}(\mbox{tp}^{\frac{1}{p}})$
with power transformation parameter $\frac{1}{p}$ and 
regression coefficients $\beta_0$, $\beta_1$, $\gamma_0$ and $\gamma_1$,  
\hlc{hereafter referred to as a basic EMOS model.}

@@

In the following, it is illustrated how distributional forests can
solve this problem without requiring prior expert knowledge.  For fitting the
forest only the response distribution and the list of potential predictor
variables need to be specified (along with a few algorithmic details) and then
the relevant variables, interactions, and potentially nonlinear effects are
determined automatically in a data-driven way. Here, we employ a zero-censored
Gaussian distribution and 80~predictor variables computed from ensemble means
and spreads of various NWP outputs. The predictive performance of the forest is
compared to three other zero-censored Gaussian models: (a)~a basic
EMOS, (b)~a GAMLSS with prespecified effects and interactions based on
meteorological knowledge/experience, and (c)~a boosted GAMLSS with automatic
selection of smooth additive terms based on all 80~predictor variables.

@@

\item \emph{EMOS:} The basic ensemble model output statistics
  models use the ensemble mean of total precipitation as regressor in the location
  submodel and the corresponding ensemble standard deviation in the scale submodel.
  The parameters are estimated by maximum likelihood, using an identity
  link for the location part and a log link for the scale part
  \citep[following the advice of][]{Gebetsberger+Messner+Mayr:2017}.

@@

\subsubsection*{East Tyrol}
Figure~\ref{fig:all-map} \hlc{shows that, contrary to the rest of Tyrol, 
for the majority of stations in East Tyrol the prespecified GAMLSS performs
best in this particular setting (learning data: 1985--2008, testing data:
2009--2012). East Tyrol is located south of the main Alpine Ridge and 
exhibits different climatological characteristics than the part north of 
the Alpine Ridge. It seems that the choice of covariates and their interactions 
we made for the prespecified GAMLSS method is particularly adequate for the 
region of East Tyrol for this specific analysis. However, the advantages of 
the prespecified GAMLSS models in this particular setting (training: 1985--2008, 
validation: 2009--2012) even out in a cross-validation framework for most of 
the stations in East Tyrol. These results are presented in Supplement 2 
(Stationwise Evaluation) where the exact same evaluations as in} 
Section~\ref{sec:axams} \hlc{for station Axams have been carried out 
for 14 other stations.}

@@

\subsubsection{Distributional specifications}
%% choice of distribution mit Verweis auf Supplement distributions
\hlc{Choosing an adequate distributional family is a crucial step and forms
the basis required to build a well fitting model. Therefore, this
decision should be made thoughtfuly. 
In the presented application we applied a Gaussian distribution left-censored 
at zero as this has proven to be a well working choice in
earlier literature} \citep{Stauffer+Mayr+Messner:2017}. 
\hlc{To test for robustness against distributional misspecifation 
different distributional assumptions have been considered as well. 
In Supplement 1 (Different Response Distributions) the exact
same evaluations as in} Section~\ref{sec:all} \hlc{are carried out for 15
observation stations, once employing a Gaussian hurdle model and once
with a logistic distribution left-censored at zero in order to account
for heavier tails. Under both assumptions the obtained results 
are similar to those presented in the paper applying a Gaussian distribution
left-censored at zero. However, comparing the two censored distributions,
at many stations the models perform not as well under the logisctic distribution 
as under the Gaussian distribution. The hurdle model leads to slight improvements 
in some settings, but this comes with more variablity (especially for EMOS and the 
boosted GAMLSS).
Alltogether, the distributional forest has proven to be very robust against
distributional misspecifications.}

\subsubsection*{Computational difficulties}
%% GAMLSS errors with references gamboostLSS Hofner et al.
As stated by \cite{Hofner+Mayr+Schmid:2016} \hlc{the AIC-based variable selection 
methods implemented in the} \textsf{R} package \textbf{gamlss} \hlc{``[...] can be 
unstable, especially when it comes to selecting possibly different sets of variables for 
multiple distribution parameters.'' We have noticed computational problems when
applying the function} \textbf{gamlss} \hlc{in certain settings within the cross-validation
framework as it did not succeed in fitting the model. In these cases the method
was not taken into consideration in the comparison of all applied models.}

@@

%\section{PIT histograms}
%\label{app:pit}

%\begin{figure}[b!]
%\centering
%%\begin{subfigure}{0.8\textwidth}
<<rain_axams_pit_oosample, fig=TRUE, echo=FALSE, width = 7, eval=FALSE>>=

@@

@
%%\end{subfigure}
%\caption{\label{fig:axams-pit}Out-of-sample PIT histograms (2009--2012) for station Axams and models learned on data from 1985--2008.}
%\end{figure}

%As an alternative to QQ~plots based on randomized quantile residuals,
%probability integral transform (PIT, \citealt{Gneiting+Balabdaoui+Raftery:2007}) 
%histograms are a commonly used tool to assess how well-calibrated distributional 
%fits are. Figure~\ref{fig:axams-pit} shows out-of-sample PIT histograms for the same
%probabilistic forecasts as Figure~\ref{fig:axams-qq}, confirming that despite
%some deviations the overall calibration of all models is acceptable.
